### 2014-06-20 paper.tex
added a major updates

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Marek's disease (MD) is caused by highly oncogenic Marek's disease virus (MDV).
Different MHC alleles have been associated with susceptibility and resistance to the disease.
However, chickens line 6 and 7 show significant phenotypic differences when chanllenged with
MDV even with the same MHC allele. Therefore, the major unanswered question is what genetic factors
underlie the susceptibility and resistance of line 6 and 7.
In this study, we identified differentially expressed genes and isoforms in response to Marek's disease infection
from chickens line 6 and 7 and performed pathway and functional analysis.
Results show that changes in expression of genes and isoforms involved in innate immune responses in line 6 may contribute
to resistance of the disease. We also identified many single nucleotide polymorphisms (SNPs) between
line 6 and 7 that may cause aberrant alternative splicing, which results in isoform with distictive functions.

@@

Marek's disease is an economically significant chicken disease that affects a poultry industry worldwide.
Total of \$2 billion loss has been reported from outbreaks~\cite{}.
The disease is caused by highly oncogenic Marek's disease virus (MDV) which causes T-cell lymphoma.
Vaccination is effective in reducing incidence of tumor formation; however, current MD vaccines do not
prevent infection or horizontal spread of the virus.
Improper use of vaccines is a key factor that drives evolution of highly virulent strains in vaccinated
flocks.

Many studies have reported strong associations between MHC alleles and resistance and susceptibility of
the disease. For example, chickens with MHC allele B$^{21}$ is highly resistant in contrast to chickens with
B$^{16}$ allele.
However, chickens line 6 and 7 , even with common allele (B$^6$), exhibit different phenotypic response to
MD.
The major unanswered question is what genetic factors contribute to susceptibility and resistance of
the disease.

Although gene expression has been shown to have a high correlation with phenotypes in many studies~\cite{}.
Studies have shown that alternative splice forms play a significant role in many biological events and
isoform expression levels can provide better signatures for some diseases~\cite{zhang2013isoform}.
Regulation of alternative splicing by enhancers (ESE) and silencers (ESS) and their corresponding
binding motifs have been identified in some organisms including human.
Mutations that disrupt or create exonic splicing enhancers could alter splicing patterns leading to aberrant alternative splicing.
In fact, 15\% of mutations that cause genetic diseases affect pre-mRNA splicing.
A number of disease-associated single-nucleotide polymorphims in coding regions (cSNPs) that affect exon splicing enhancers (ESEs) and
silencers (ESSs) have been characterized~\cite{blencowe2000exonic, wang2007splicing}.
We hypothesized that SNPs between line 6 and 7, which are located in exons, are responsible for differential
expression of some isoforms between chicken line 6 and line 7.
We reported some genes that could be regulated by \textit{cis-}regulatory elements that may also play a role in
immune reponses.

% Results and Discussion can be combined.
\section*{Results and discussion}

@@

Many genes were found to be differentially expressed (DE) between control and infected
chickens in line 6 and line 7.
While the number of unique downregulated genes in both lines are approximately equal,
the number of unique upregulated genes in line 7 is much greater than in line 6 as shown
in figure~\ref{degenes_venn}.
Interestingly, some genes that were differentially expressed in both lines were
regulated in the opposite direction (table~\ref{tab:opposite}).
Among genes downregulated in line 6 but upregulated in line 7,
\textit{LL} (lung lectin) and \textit{SFTPA1}, which encode lectin proteins, are part of
lectin-complement pathway.
\textit{LIMS1} is involved in cell differentiation and proliferation and
\textit{PPARG} is a suppressor of \textit{NF$\kappa$B}-mediated proinflamatory response.
On the other hand, nearly all genes upregulated in line 6 but downregulated in line 7
are involved in cell survival such as mRNAs splicing, cell growth, and protein systhesis,
except CD7 whose function is involved in T cell-B cell interaction.
This difference suggests that even at this stage of infection in line 6, the lytic phase 
could be controlled or subsided because the viruses are driven to undergo latent infection early.
Therefore, genes important for recovery of cell damages are upregulated.
In comparison, the lytic phase in line 7 may still continue and as a result, genes involved in
immune responses are still upregulated.
In addition, type I interferon (\textit{IFN-$\gamma$} and \textit{IFN-$\beta$}) as well as \textit{INF-$\alpha$3} were found to be highly
upregulated in infected chickens in both line 6 and 7 (table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors
were not different in line 6, but upregulated in line 7.
This could also reflect the ongoing immune responses in line 7.

Some DE genes have been reported from previous Microarray studies were also found to be
differentially expressed in this study.
For example, B6.1 (Bu-1) gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with MHC allele B\textsuperscript{19} at 4 d.p.i~\cite{sarson2008transcriptional}.
It was also found to be down-regulated about 3-fold in line 7 (susceptible).
Similary, \textit{GMZA} gene reported to be upregulated across genetically different chickens
(B$^{19}$, B$^{21}$ alleles), was also found to be highly upregulated.
In contrast, some genes that have been reported to be highly expressed in resistant chickens were
downregulated in both lines. Those genes are \textit{AMIGO2, MMP13} and \textit{CLEC3B}, which were
found to be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed in resistant chickens:
% \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line 7$_{2}$) including
\textit{AVD, ART1, NOS2, CXCL13L2, MX1} and \textit{SOCS1}~\cite{smith2011systems}
were also found to be highly upregulated in both line 6 and 7.
However, our results show the similar expression pattern of \textit{IL6} and \textit{IL18}, which were
only upregulated in line 7 in early stage of infection (3-5 d.p.i).

In contrary, \textit{IL15} has been reported non differential-expressed between control and infected chickens in
line 6$_{1}$ and 7$_{2}$~\cite{kaiser2003differential}; however,
it was only upregulated in line 7 in our study.
Expression of \textit{IL15} is induced by \textit{TLR9}, which binds to non-methylated CpG
residues present in genomes of many DNA viruses, including herpes simplex virus.
This cytokine auto-regulates the expression of \textit{CD40}, which is
a transmembrane receptor required for activation of macrophages by CD4 T cells.
Consequently, \textit{CD40} was only upregulated in line 7 (data not shown).

\subsection*{Functional analysis of differntial-expressed genes}

To determine pathways that were perturbed during the infection, data were
analysed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq data~\cite{young2010method}.
Significantly perturbed pathways (FDR $< 0.1$) from both lines that involved in
immune response include Toll-like receptor signaling pathway, cytokine-cytokine receptor interaction,
intestinal immune network for IgA production and cell adhesion molecules (CAMs).
Some other pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling pathway,
NOD-like receptor signaling pathway and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed genes in phagosome pathway.
Although MHC I gene (\textit{BF1}) was differentially expressed in both lines, other genes involved in
expressing newly synthesized MHC I were only upregulated in line 7
suggesting that new MHC I molecules were actively produced and expressed on cell surface.
Furthermore, gene ontology analysis of biological processes (GO:BP) shows that categories involved in
both adaptive and innate immune responses were enriched in line 7 (supplementary material).
On the other hand, only categories involved in innate immune responses were enriched in line 6.
In addition, enrichment of apoptosis pathway in line 7 indicates that the programmed
cell death could be induced by CTL response to eliminate the viruses.

At this stage of infection, our results suggest that lytic infection of MDV stimulates
both innate and the adaptive immune response, which leads to activation of T cells in line 7.
Only activated T cells are infected by MDV, therefore, the lytic phase could facilitate the spread of
the viruses by enhancing expansion of activated T cells.
In contrast, the immune response in line 6 could be able to drive the viruses to latency phase early
and only activate the adaptive immune response in a very short duration.
Due to cell-associate nature of MDV, the viruses transfer to T cells via cell-to-cell contact between B cells and
T cells during antigen presentation or B cells activation by T helper cells.
Therefore, it is beneficial for the host to restrain such contact.
However, it is not clear how chickens in line 6 control the lytic infection of MDV.
% Two mechanisms have been speculated to contribute to MD resistance.
% First innate immune responses could be highly effective and could activate
% strong adaptive immune responses that rapidly control viral replication and force the viruses to undergo latent phase.
% Second, innate immune resposnes itself could be highly effective in limiting viral replication~\cite{smith2011systems}.

@@

Immune system is isoform-rich and many genes express different isoforms with distinctive functions in response to
stimuli such as stress, chemicals and infection.
Changes in expression of splice forms of immune related genes have been reported to be associated with increase
susceptibility and poor prognosis of diseases~\cite{lynch2004consequences}.
Studying differential isoform expression could therefore shed lights into inherent differences between 
line 6 and 7 that confer resistance or susceptibility to MD.

In the past decades, microarray technology has been used to study gene and isoform expressions in many studies, but
its sensitiviy for detection of structurally similar isoforms is low and known or predicted annotations
are required to design probes~\cite{kane2000assessment}.
Although RNA-Seq method can provide a reliable estimate of an exon expression compared to microarray~\cite{pan2008deep}
and is not constrained to those limitations, studying isoform
expressions using RNA-Seq is still not straightforward because of short read lengths.
Reads from current Illumina technology are generally not long enough to span across all exons in an isoform.
In most cases, only exons in close proximity are covered by the same read, which makes it
difficult to accurately predict a full structure of the isoform.
In addition, some genes are fused due to overlapping untranslated regions (UTRs),
which can also result in erroneous predicted isoform structures.

Due to those issues, it is not feasible to accurately estimate exression of isoforms, especially when
gene annotation is constructed from \textit{de novo} assembly~\cite{trapnell2013differential}.
To avoid these issues, we chose to study exon expression instead of isoform expression.
Using MISO with exon-centric method, only reads spanning across a few exons are used and only exons
involved in a splicing event are examined.
The expression of exon inclusion is calculated as percent spliced in (Psi or $\Psi$), which can be used to
infer the portion of transcripts that include the exon in each sample~\cite{Katz:2010iv}.
In this study, we investigated the three most common alternative splicing events in vertebrates, which are skipped
exons (SE), an alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.

@@

List of DEU genes from line 6 that show difference in $\Psi$ greater than 0.20 when compared to line 7
in infected chickens is shown in table~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.
Genes can be categorized roughly into four groups based on the pattern of $\Psi$ across control and
infected birds in both lines.

Group I includes genes that $\Psi$s were up or down regulated in infected chickens in line 6 only.
This group include \textit{BCL11B} (B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell lymphoma
associated C2H2-type zinc finger protein encoding gene, which functions as a tumour-suppressor
in T-cell lymphoma in human.
According to homologous alignments on UCSC genome browser, a splice form with the skipped exon
is similar to mouse \textit{BCL11B isoform b}.
The skipped exon was expressed 30\% in infected line 6 chickens; whereas it was rarely expressed
(4-7\%) in the control line 6 and both groups in line 7.
The skipped exon was not found to encode any known protein domain, however, it is in the middle of
two adjacent C2H2-type finger protein domains.
Another gene related to B and T cell lymphoma is \textit{SIK2} (salt-inducible kinase 2).
This gene has been reported to have a negative effect on T cell lymphoma by limiting the transcription of
HTLV-1 virus~\cite{tang2013lkb1}.
Some other genes are important in pre-mRNA splicing including \textit{TRA2A}, \textit{SRSF6} and \textit{GEMIN6}.
\textit{TRA2A} (transformer-2 protein homolog alpha) encodes RNA recognition motif (RRM),
the skipped exon is not found to encode any known protein domain.
\textit{SRSF6} (SR splicing factor 6) encodes a nuclear protein that belongs to the splicing factor protein family.
\textit{GEMIN6} plays a role in the assembly of spliceosomal snRNP in cytoplasm.
These genes could play a significant role in regulating inclusion of alternative exons.
A gene in this group that might be important for innate immune responses is \textit{RAC3} (Ras-related C3
botulinum toxin substrate 3).
This gene encodes small GTPases, belonging to Ras family, that regulates a wide variety of cellular events
including cell growth, cytoskeletal reorganization and the activation of protein kinases.
The role of small GTPases in immune responses is discussed further below.

In group II, $\Psi$ values were relatively stable in control and infected chickens in line 6 and line 7.
Genes that could play an important role in immune responses are \textit{ITGB2} and \textit{HCK}.
\textit{ITGB2} (CD18) encodes subunit $\beta_{2}$ integrin of \textit{LFA-1} and \textit{CR3} receptors.
\textit{LFA-1} plays an important role in adhesion of lymphocytes with other cells.
\textit{CR3} binds to a vast array of ligands and molecules including complement C3bi, microbial proteins,
ICAM-1 and -2, ECM proteins and coagulation proteins~\cite{}.
It plays a significant role in neutrophils and monocytes activation including phagocytosis, adhesion and
migration~\cite{}.
Mutations in CD18 gene has been associated with type 1 leukocyte adhesion deficiency (LAD-1),
an autosomal-recessive inherited disease found in few families.
The disease is characterized by impair of lymphocytes in adherent-dependent functions, lack of accumulation
to the site of infection and recurrent bacterial and fungal infection~\cite{}.
\textit{HCK} transmits signal from cell surface receptors such as \textit{FCGR1A, FCGR2A, IL2, IL6,
IL18} and integrins (\textit{ITGB1, ITGB2}).
Based on our gene model, the skipped exon of \textit{ITGB2} is located in the 5$\prime$ UTR;
whereas, the skipped exon of \textit{HCK}, encodes protein tyrosine kinase (Pfam:PF07714).
Some other genes in this group are involved in cell rearrangement and cytokinesis that could be
regulated by signals from integrins including \textit{DYLNT1,DYNLL2, SEPT11, PFN2} and \textit{ZDHHC7}.
The role of \textit{ITGB2} in innate immune responses will be discuss further in the next section.

Group III includes genes that exhibit differential isoform expression only in infected chickens in line 7.
A number of genes in this group encode proteins that are parts of spliceosome: \textit{SRSF3}, \textit{HNRNPDL},
\textit{SFSWAP}, \textit{THOC1}, \textit{RNPC3} and \textit{SRSF5}.
Two genes involved in cell-cell contact regulated by integrins are also in this group.
\textit{PPP1R12A} is a myosin phosphatase that regulates the interaction of actin and myosin downstream of GTPase
Rho proteins.
\textit{PODXL} encodes protein that functions in integrin-dependent manner to increase cell-cell contact.
Connections of these two genes and integrins will be discussed in the next section.

The last group only has one gene, \textit{GOSR1}.
The $\Psi$ value of this gene was less than 0.20 cutoff in control and infected chickens in line 6, but it is
significantly different between infected chickens in line 6 and 7.
Also, there is a significant difference between control and infected chickens in line 7.
This gene encodes traffiking membrane protein important for transporting proteins from \textit{cis-} to
\textit{trans-}golgi network.

\subsubsection*{Integrins and actin cytoskeleton pathway}

Integrins are a special kind of receptors that transmit signals bidirectionally across cell membrane.
They are heterodimeric composed of an $\alpha$ (large) and a $\beta$ (small) subunits.
$\beta_{2}$ (CD18) subunit is exclusively expressed on lymphocytes and APCs as a component of \textit{LFA-1}
and \textit{CR3} receptors.

(more on \textit{LFA-1} and \textit{CR3} to be added).

As discussed above, \textit{RAC3, HCK, DYLNT1, DYNLL2, SEPT11, PFN2, ZDHHC7, PODXL} and \textit{PPP1R12A}
could be activated by the integrins because of their roles in cell rearrangement or molecular interaction.
Interestingly, we found that in human, \textit{RAC3, ITGB2, PFN2} and \textit{PPP1R12A} are in the regulation
of actin cytoskeleton pathway (figure~\ref{kegg_actin}).
Moreover, at least two of these genes are co-present in other three pathways (table~\ref{tab:integrin}) that
involved in immune responses.
This strongly suggests that these four genes could contribute to difference in immune responses between
line 6 and 7 even though their expressions were not significantly different between control and infected groups.

% \subsubsection*{Splicing factors}
% 
% (Not sure if I should write about splicing factors, I'm afraid it will make the paper too long and unfocused.)
% 
% The second group includes genes encode protein compoments of spliceosome or are involved in pre-mRNA splicing.
% Those genes are \textit{TRA2A, SRSF6, GEMIN6, SRSF3, HNRNPDL, SFSWAP, THOC1, THOC2, RNPC3} and \textit{SRSF5}.
% Even though alternative splicing is regulated by splicesome, it is difficult to elucidate the effect of alternative
% splice forms of spliceosome subunits on alternative splicing regulation.
% This is in part due to the fact that the spliceosome is the most complex protein structures composed of many different
% subunits each of which could have isoforms differing in function or activity~\cite{}.

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Marek's disease (MD) is caused by highly oncogenic Marek's disease virus (MDV).
Understanding the genetic basis for non-MHC resistance would be of both fundamental and applied importance.
In this study, using computational approaches of RNA-Seq data, we identified
differentially expressed genes and isoforms in response to MDV infection
from highly inbred chickens lines that are MD resistant and susceptible,
and performed pathway and functional analysis.
Results show that expression of genes and isoforms involved in cell cytoskeleton and
adhesion molecules in line 6 may contribute to resistance of the disease by limiting
T cell activation, which in turn prevents the spread of MDV to activated T cells.
In addition, we identified exonic single nucleotide polymorphisms (SNPs) between
lines 6 and 7 predicted their role in regulation of alternative splicing.
Based on our results, we speculate that alternative splice forms of cell adhesion molecules
and cytoskeleton play and important role in controlling T cell activation in line 6.

@@

Marek's disease (MD) is an economically significant chicken disease that affects the
poultry industry worldwide with estimated annual loss of
\$2 billion loss has been reported from outbreaks~\cite{morrow2004marek}.
The disease is caused by the highly oncogenic Marek's disease virus (MDV)
an alphaherpesvirus that induces T-cell lymphomas in susceptible birds.
Vaccination is the primary control measure, which is effective in reducing
incidence of tumor formation.
However, since MD vaccines are not sterilizing, they do not prevent infection or
horizontal spread of the virus.
As a consequence, MDV field strains that overcome vaccinal protection have arisen repeatedly over time.
Therefore, there is a need for sustainable alternative controls measures, such as improving genetic
resistance.

Many studies have reported strong associations between MHC alleles and resistance or susceptibility to
MD. For example, chickens with MHC allele B$^{21}$ are highly resistant in contrast to chickens with
B$^{16}$ allele which are highly susceptible.
ADOL lines 6 and 7, both share (B$^6$) allele, yet exhibit different phenotypic responses to
MD; e.g., challenge with JM/102W strain with result in 0 and 100\% MD incidence for lines 6 and 7,
respectively.
Thus, the major unanswered questions are what genetic factors, especially those that are non-MHC,
contribute to susceptibility and resistance of the disease and what is the main contributing
mechanism.

In the past decades, significant efforts have been spent to study variations in a global gene
expression between resistant and susceptible birds using
microarray and RNA-Seq methods in order to identify non-MHC genes that contribute
to resistance to MD~\cite{}.
However, none of the studies have investigated differential expression of alternative isoforms,
which are known to play a significant role in many biological events including
immune responses.
In addition, studies have shown that isoform expression levels can provide better signatures for some
diseases~\cite{zhang2013isoform}.
Changes of isoform expression levels are governed partly by two types of \textit{cis-}regulatory
elements: Exon Splicing Enhancer (ESE) and Exon Splicing Silencer (ESS),
which are located within an exon sequence.
A number of sequence motifs of ESE and ESS have been identified in human and some other organisms
and could be predicted \textit{in silico}.
Mutations that disrupt or create those motifs could alter splicing
patterns leading to aberrant alternative splicing.
A number of disease-associated single-nucleotide polymorphsims in coding regions
(cSNPs) that affect ESEs and ESSs have been well
characterized~\cite{blencowe2000exonic, wang2007splicing}.
%In fact, 15\% of mutations that cause genetic diseases affect pre-mRNA splicing via this mechanism.
Therefore, variations in isoform expression could lead to identification of SNPs
that underlie genetic resistance to MD.
In this article, we reported differential-expressed genes and isoforms that may contribute to
resistance to MD as well as cSNPs that can potentially affect isoform expression levels.

% Results and Discussion can be combined.
\section*{Results}

@@

Many genes were found to be differentially expressed (DE) between control and infected
chickens in lines 6 and line 7.
While the number of unique downregulated genes in both lines was approximately equal,
the number of unique upregulated genes in line 7 was much greater compared
to line 6 (Figure~\ref{degenes_venn})~\cite{lee1983ontogeny}.
Interestingly, some genes that were differentially expressed in both lines were
regulated in the opposite direction (Table~\ref{tab:opposite}).
Among genes downregulated in line 6 but upregulated in line 7 were
\textit{LL} (lung lectin) and \textit{SFTPA1}, which encode a calcium-dependent C-type
lectin and a lung surfactant protein respectively.
Both molecules are important in innate immunity~\cite{hogenkamp2008chicken,kingma2006defense}.
\textit{LIMS1} is involved in cell differentiation and proliferation and
\textit{PPARG} is a suppressor of \textit{NF$\kappa$B}-mediated proinflamatory response.
On the other hand, nearly all genes upregulated in line 6 but downregulated in line 7
are involved in cell survival such as mRNAs splicing, cell growth, and protein synthesis,
except CD7 whose function is involved in T cell-B cell interaction.
This difference suggests that even at this stage of infection in line 6, the lytic phase 
could be subsided, therefore, only genes important for survival of cell are upregulated in line 6.
In comparison, the lytic phase in line 7 may still continue and as a result, genes involved in
immune responses are still upregulated.
In addition, type I interferon (\textit{IFN-$\gamma$} and \textit{IFN-$\beta$}) as well as \textit{INF-$\alpha$3} were found to be highly
upregulated in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors
were not different in line 6, but upregulated in line 7.
This could also reflect the ongoing immune responses in line 7.

Some DE genes have been reported from previous microarray studies were also found to be
differentially expressed in this study.
For example, B6.1 (Bu-1) gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4 d.p.i~\cite{sarson2008transcriptional}.
It was also found to be down-regulated about 3-fold in line 7 (susceptible).
Similarly, \textit{GMZA} gene reported to be upregulated across genetically different chickens
(B$^{19}$, B$^{21}$ alleles), was also found to be highly upregulated.
In contrast, some genes that have been reported to be highly expressed in resistant chickens were
downregulated in both lines. Those genes are \textit{AMIGO2, MMP13} and \textit{CLEC3B}, which were
found to be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed in resistant chickens:
% \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line 7$_{2}$) including
\textit{AVD, ART1, NOS2, CXCL13L2, MX1} and \textit{SOCS1}~\cite{smith2011systems}
were also found to be highly upregulated in both line 6 and 7.
However, our results show the similar expression patterns for \textit{IL6} and \textit{IL18}, which were
only upregulated in line 7 in early stage of infection (3-5 d.p.i).

In contrast, \textit{IL15} has been reported non differential-expressed between control
and infected chickens in
line 6$_{1}$ and 7$_{2}$~\cite{kaiser2003differential}; however,
it was only upregulated in line 7 in our study.
Expression of \textit{IL15} is induced by \textit{TLR9}, which binds to non-methylated CpG
residues present in genomes of many DNA viruses, including herpes simplex virus.
This cytokine auto-regulates the expression of \textit{CD40}, which is
a transmembrane receptor required for activation of macrophages by CD4 T cells.
Consequently, \textit{CD40} was only upregulated in line 7 (data not shown).

\subsection*{Functional analysis of differential-expressed genes}

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq data~\cite{young2010method}.
Significantly perturbed pathways (FDR $< 0.1$) from both lines that involved in
immune response include TLR signaling pathway, cytokine-cytokine receptor interaction,
intestinal immune network for IgA production, and cell adhesion molecules (CAMs)~\ref{line67_kegg}.
Some other pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling pathway,
NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.
Although MHC I gene (\textit{BF1}) was differentially expressed in both lines, other genes involved in
expressing newly synthesized MHC I were only upregulated in line 7
suggesting that new MHC I molecules were actively produced and expressed on cell surface.
% (the minor class I is upregulated but the major class I is downregulated)
Furthermore, gene ontology analysis of biological processes (GO:BP) shows that categories involved in
both adaptive and innate immune responses were enriched in line 7 (supplementary material).
On the other hand, only categories involved in innate immune responses were enriched in line 6.
In addition, enrichment of apoptosis pathway in line 7 indicates that the programmed
cell death could be induced by CTL response to eliminate the viruses.

At this stage of infection, our results suggest that lytic infection of MDV stimulates
both innate and adaptive immune responses, which leads to activation of T cells in line 7.
Only activated T cells are infected by MDV, therefore, the lytic phase could facilitate the spread of
the viruses by enhancing expansion of activated T cells.
Due to cell-associate nature of MDV, the viruses transfer to T cells via cell-to-cell
contact between B cells and
T cells during antigen presentation or B cells activation by T helper cells.
Therefore, it is beneficial for the host to restrain such contact.
However, it is not clear how chickens in line 6 control the lytic infection of MDV.
% Two mechanisms have been speculated to contribute to MD resistance.
% First innate immune responses could be highly effective and could activate
% strong adaptive immune responses that rapidly control viral replication and force the viruses to undergo latent phase.
% Second, innate immune resposnes itself could be highly effective in limiting viral replication~\cite{smith2011systems}.

@@

The immune system is isoform-rich and many genes express different isoforms with
distinctive functions in response to stimuli such as stress, chemicals and infection.
Changes in expression of splice forms of immune related genes have been reported to
be associated with increase susceptibility and poor prognosis of diseases~\cite{lynch2004consequences}.
Studying differential isoform expression could therefore shed lights into inherent differences between 
lines 6 and 7 that confer resistance or susceptibility to MD.

In the past decades, microarray technology has been used to study gene and isoform
expressions in many studies, but its sensitivity for detection of structurally similar
isoforms is low, and known or predicted annotations are required to design probes~\cite{kane2000assessment}.
Although RNA-Seq method can provide a reliable estimate of an exon expression compared
to microarray~\cite{pan2008deep} and is not constrained to those limitations, studying isoform
expressions using RNA-Seq is still not straightforward because of the short read lengths.
Reads from current Illumina technology are generally not long enough to span across all exons in an isoform.
In most cases, only exons in close proximity are covered by the same read, which makes it
difficult to accurately predict a full structure of the isoform.
In addition, some genes are fused due to overlapping untranslated regions (UTRs),
which can also result in erroneous predicted isoform structures.

Due to those issues, it is not feasible to accurately estimate expression of isoforms, especially when
gene annotation is constructed from \textit{de novo} assembly~\cite{trapnell2013differential}.
To avoid these issues, we chose to study exon expression instead of isoform expression.
Using MISO with exon-centric method, only reads spanning across a few exons are used and only exons
involved in a splicing event are examined.
The expression of exon inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used to
infer the portion of transcripts that include the exon in each sample~\cite{Katz:2010iv}.
In this study, we investigated the three most common alternative splicing events in
vertebrates, which are skipped exons (SE), an alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.

@@

Lists of DEU genes from line 6 that show difference in $\Psi$ greater than 0.20 when compared to line 7
in infected chickens are shown in
Tables~\ref{tab:line67i_diff_line67u_one},~\ref{tab:line67i_diff_line67u_two}
and~\ref{tab:line67i_diff_line67u_three}.
Genes can be categorized roughly into four groups based on the pattern of $\Psi$ across control and
infected birds in both lines.

Group I (Table~\ref{tab:line67i_diff_line67u_one}) includes genes with $\Psi$s that were
up- or down-regulated in infected chickens in line 6 only.
This group includes \textit{BCL11B} (B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell lymphoma
associated C2H2-type zinc finger protein encoding gene, which functions as a tumour-suppressor
in T-cell lymphoma in human.
According to homologous alignments on UCSC genome browser, a splice form with the skipped exon
is similar to mouse \textit{BCL11B isoform b}.
The skipped exon was expressed 30\% in infected line 6 chickens; whereas it was rarely expressed
(4-7\%) in the control line 6 and both groups in line 7.
The skipped exon was not found to encode any known protein domain, however, it is in the middle of
two adjacent C2H2-type finger protein domains.
Another gene related to B and T cell lymphomas is \textit{SIK2} (salt-inducible kinase 2).
This gene has been reported to have a negative effect on T cell lymphomas by limiting the transcription of
HTLV-1 virus~\cite{tang2013lkb1}.
Some other genes are important in pre-mRNA splicing including \textit{TRA2A},
\textit{SRSF6} and \textit{GEMIN6}.
\textit{TRA2A} (transformer-2 protein homolog alpha) encodes RNA recognition motif (RRM),
the skipped exon is not found to encode any known protein domain.
\textit{SRSF6} (SR splicing factor 6) encodes a nuclear protein that belongs to the splicing
factor protein family.
\textit{GEMIN6} plays a role in the assembly of spliceosomal snRNP in cytoplasm.
These genes could play a significant role in regulating inclusion of alternative exons.
A gene in this group that might be important for innate immune responses is \textit{RAC3} (Ras-related C3
botulinum toxin substrate 3).
This gene encodes small GTPases, belonging to Ras family, that regulates a wide variety of cellular events
including cell growth, cytoskeletal reorganization, and the activation of protein kinases.
The role of small GTPases in immune responses is discussed further below.

In group II (Table~\ref{tab:line67i_diff_line67u_two}),
 $\Psi$ values were relatively stable in control and infected chickens within line, but not between lines.
Genes that could play an important role in immune responses are \textit{ITGB2} and \textit{HCK}.
\textit{HCK} transmits signal from cell surface receptors such as \textit{FCGR1A, FCGR2A, IL2, IL6,
IL18}, and integrins (\textit{ITGB1, ITGB2}).
\textit{ITGB2} (CD18) encodes subunit $\beta_{2}$ integrin of \textit{LFA-1} and \textit{CR3} receptors.
\textit{LFA-1} plays an important role in adhesion of lymphocytes with other cells.
\textit{CR3} binds to a vast array of ligands and molecules including complement C3bi, microbial proteins,
ICAM-1 and -2, ECM proteins, and coagulation proteins~\cite{}.
It plays a significant role in neutrophils and monocytes activation including phagocytosis, adhesion and
migration~\cite{}.
The role of \textit{ITGB2} in immune responses is discussed further in the next section.
\textit{DYNLT1,DYNLL2, SEPT11, PFN2}, and \textit{ZDHHC7} are also involved in cell
rearrangement and cytokinesis.
In particular, \textit{DYNLT1} and \textit{DYNLL2} are dynein proteins that have been demonstrated to
regulate T cell activation by driving T cell receptor microclusters (TCR-MCs)
toward the center of immune synapse~\cite{hashimoto2011dynein}.

Group III (Table~\ref{tab:line67i_diff_line67u_three})
includes genes that exhibit differential isoform expression only in infected line 7.
A number of genes in this group encode proteins that are parts of
spliceosome: \textit{SRSF3}, \textit{HNRNPDL},
\textit{SFSWAP}, \textit{THOC1}, \textit{RNPC3} and \textit{SRSF5}.
Two genes involved in cell-cell contact regulated by integrins are also in this group.
\textit{PPP1R12A} is a myosin phosphatase that regulates the interaction of actin
and myosin downstream of GTPase Rho proteins.
\textit{PODXL} encodes PODX-like protein that functions in integrin-dependent
manner as both pro-adhesive and anti-adhesive molecules.
This protein is involved in cell-to-cell contact, cell trafficking, and cancer progression~\cite{}.

The last group (Group IV, Table~\ref{tab:line67i_diff_line67u_three})
only has one gene, \textit{GOSR1}.
The $\Psi$ value of this gene was less than 0.20 cutoff in control and infected chickens in line 6, but it is
significantly different between infected chickens in line 6 and 7.
Also, there is a significant difference between control and infected chickens in line 7.
This gene encodes a trafficking membrane protein important for transporting proteins
from \textit{cis-} to \textit{trans-}golgi network.

\subsubsection*{Roles of \textit{LFA-1} and actin cytoskeleton pathway in T cells activation}

Integrins are a special kind of receptors that transmit signals bidirectionally across the cell membrane.
They are heterodimeric composed of an $\alpha$ (large) and a $\beta$ (small)
subunits~\cite{wang2010immunopathologies}.
$\beta_{2}$ (CD18) subunit encoded by \textit{ITGB2} is expressed on lymphocytes and antigen presenting cells
(APCs) as a component of \textit{LFA-1} and \textit{CR3} receptors.
LFA-1 binds to its ligand ICAM-1 to help form a synapse that bring APC and T cell together to
initiate antigen presentation leading to T cell activation~\cite{dustin2000immunological}.
Absence of LFA-1 leads to impaired functions of lymphocytes in proliferation and
tumor rejection~\cite{scharffetter1998spontaneous,schmits1996lfa}.
Mutations in \textit{ITGB2} gene has been associated with type 1 leukocyte adhesion deficiency (LAD-1),
an autosomal-recessive inherited disease found in few families.
The disease is characterized by impair of lymphocytes in adherent-dependent functions, lack of accumulation
to the site of infection and recurrent bacterial and fungal infection~\cite{springer1987lymphocyte}.

Actin cytoskeleton is very important in T cell activation because it enhances activity of
LFA-1 by increasing its avidity and recruiting signaling molecules necessary for
downstream signaling~\cite{dustin2000immunological, van2000avidity}.
Cytoskeleton proteins binding to cytoplasmic domain of LFA-1 are thought to play an important role in
driving LFA-1 to aggregation on the cell surface, resulting in increased avidity.
Aggregation of LFA-1 has been demonstrated to be essential for lymphocytes to bind to the
ligand~\cite{van1994extracellular}.
Interestingly, \textit{RAC3}, \textit{PFN2}, and \textit{PPP1R12A}, which are involved in
actin cytoskeleton pathway (Figure~\ref{kegg_actin}), also expressed different ratios of alternative splice
forms between lines 6 and 7.
Some of these genes are also co-present in three other pathways
that involved in immune responses (Table~\ref{tab:integrin}).
This could be speculated that pre-mRNA splicing of these genes is co-regulated by splicing
regulators or some genetic factors.
